---
title: "Programming for Economists"
author: "Alexander Tijburg.2673913., Martijn Lustig.2781885., Duco Mol .2833134., Andreea Muntean, Seerp van Roijen .2743665. and Ikhouane Imran"
Tutorial_group: Late_Group2
output: pdf_document
date: "2025-06-14"
---

# Environment setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(dplyr)
library(ggplot2)
library(jsonlite)
```

# introduction and research relevance

When COVID-19 struck the global economic system, journalists, analysts, and company officials were quick to point out the significant effects this crisis had on unemployment numbers. Contextually referring specifically to the portion of the labor force that is willing and able to work but unable to find work. Headlines of articles addressing this topic compared the dramatic rise in unemployment to that of the Great Recession,^1 whilst other, more down-to-earth tones could not help but to lament the jobs that would have been created without this intervention in the economy.^2
However, whilst the unemployment levels are rightfully a concern to governments and policymakers, this concern should not undermine the attention given to other marginalized socioeconomic demographics. One such group are the working poor, people who are working, and thus a part of the productive workforce within society, and who simultaneously are below the poverty line. A report from Eurofund (the European Foundation for the improvement of Living and Working Conditions) on the working poor concludes that in the aftermath of the 2008-2009 Financial Crisis, unemployment was seen as a priority, whilst the working poor were not.^3 
	Being a member of the working poor is associated with a range of negative consequences on the quality of life of an individual. The financial uncertainty and instability of being in such a situation brings with it heightened levels of stress, which in turn weakens a person's immune system throughout the long term. Furthermore these conditions  contribute to a higher vulnerability of mental health issues among members of the working poor.^4 The ability to smooth economic shocks through insurance can lead to a reduction of stress, a mechanism of our current economic system that is not available to many of the working poor.^5 In large-scale social surveys assessing the quality of life of the working poor, the group consistently reports significantly lower satisfaction relative to other socioeconomic groups. Additionally a higher percentage of respondents also indicated that they experience social seclusion.^6 The working poor are by no means a small group; in 2017, this entailed 9.4% of those working in the EU and 6.1% in the Netherlands.^7 
	Since the onset of the pandemic in 2020, many countries were forced into lockdown, this was accompanied by a sharp decline of economic activity which further exacerbated the already existing socio economic problems. If our society is to address the issue of the working poor in the future, which might include another pandemic, then we need to prepare policies that can help these people out. However, in order to begin policymaking in earnest the government, its municipalities and the relevant involved institutions need analyses of data on the working poor to be able to plan for both the long-term scenario and scenarios in which economic activity suddenly slows down. To that end this research report aims to conduct a data analysis of the working poor in the Netherlands. This report will assess the fluctuations of the quantities of working poor for the spatial variable of city sizes and for the temporal variable of the pre and post-COVID societies.

Data Sourcing and Description
Our team opted to use data provided by the Centraal Bureau voor Statistiek. This institution answers to the Dutch government, which is one of the earliest countries in many categories of data collection, and has as its sole objective to collect reliable and coherent data to be used by society. The CBS provides a wealth of data through their electronic free-to-use database Statline. The website allows researchers to look by themes; however, there is also a search bar. After going through the motions of checking some datasets, our team settled on: Laag en langdurig laag inkomen; huishoudenskenmerken, regio (indeling 2024), and on Bevolking; herkomstland, geboorteland, leeftijd, regio.
	The first dataset we chose was according to the assignment’s need for a social problem. The second dataset was chosen parallel to the formulation of our research question; We decided to look at the working-poor within settlements of varying sizes over a timespan that covered the Covid-19 pandemic. The dataset on the long-term poor does have some information on the population structure, but our second dataset goes into much more depth, which allows us the freedom to come up with more interesting variables. 
	The metadata of the datasets gives an explanation of many variables, which makes it easier to think of concepts of variables that can be created; however, many specific sub-population variables are not relevant to our current research. The metadata does specify the years for which data is present. These are 2018, 2019, 2022, and 2023. The dataset skips 2021 and 2022; entries for these years would have been welcome, but they do not matter for our overall research project, as we will be comparing pre- and post-COVID society. 



Footnotes: 
1 Eurofound. (2017). In-work poverty in the EU. Publications Office of the European Union, 27. https://doi.org/10.2806/65550 
2 Gould, E., & Kassa, M. (2021, May 20). Low-wage, low-hours workers were hit hardest in the COVID-19 recession: The State of Working America 2020 employment report. Economic Policy Institute. https://www.epi.org/publication/swa-2020-employment-report/
3 Eurofound. (2017). In-work poverty in the EU. Publications Office of the European Union, 35. https://doi.org/10.2806/65550 
4  Patel, J. A., Nielsen, F. B. H., Badiani, A. A., Assi, S., Unadkat, V. A., Patel, B., Ravindrane, R., & Wardle, H. (2020). Poverty, inequality and COVID-19: The forgotten vulnerable [Letter to the editor]. Public Health, 183, 110–111. https://doi.org/10.1016/j.puhe.2020.05.006
5 Ridley, M., Rao, G., Schilbach, F., & Patel, V. (2020). Poverty, depression, and anxiety: Causal evidence and mechanisms. Science, 370(6522), eaay0214. https://doi.org/10.1126/science.aay0214 
6 Eurofound. (2017). In-work poverty in the EU. Publications Office of the European Union, 27. https://doi.org/10.2806/65550 
7 Eurostat. (2019). Living and working in Europe 2019: Statistical pocketbook. Publications Office of the European Union, 4. https://ec.europa.eu/eurostat/documents/3217494/10335481/KS-02-19-303-EN-N.pdf
8 See Mandemakers, K. (2000). The Netherlands Historical Sample of the Netherlands (HSN). In P. Kelly Hall, R. McCaa, & G. Thorvaldsen (Eds.), Handbook of International Historical Microdata for Population Research, 157. Minnesota Population Center.
9 Centraal Bureau voor de Statistiek. (2024). Laag en langdurig laag inkomen; huishoudenskenmerken, regio (indeling 2024) [Data set]. CBS StatLine. https://opendata.cbs.nl/statline/portal.html?_catalog=CBS&_la=nl&tableId=86006NED&_theme=310
10 Centraal Bureau voor de Statistiek. (2024). Bevolking; herkomstland, geboorteland, leeftijd, regio, 1 januari [Data set]. CBS StatLine. https://opendata.cbs.nl/statline/portal.html?_catalog=CBS&_la=nl&tableId=85458NED&_theme=278



```{r}
Laag_Inkomen <- read.csv("Lage Inkomens.csv", sep = ";")
```
```{r}
 Bevolking <- read.csv("Bevolking.csv", sep = ";")

```
```{r}
# Merge Laag_Inkomen met Bevolking op RegioS en Perioden
Laag_Inkomen_merged <- merge(
  Laag_Inkomen,
  Bevolking[, c("RegioS", "Perioden", "BevolkingOp1Januari_1")],
  by = c("RegioS", "Perioden"),
  all.x = TRUE  # Zorgt dat alle rijen van Laag_Inkomen behouden blijven
)

```


```{r}
library(dplyr)

Laag_Inkomen_merged <- Laag_Inkomen_merged %>%
  mutate(Totaal_werkende_armen = (Personen_2 / 1000) * BevolkingOp1Januari_1)
#rounding
Laag_Inkomen_merged <- Laag_Inkomen_merged %>%
  mutate(Totaal_werkende_armen = round(Totaal_werkende_armen))

```
```{r}
library(dplyr)

Laag_Inkomen_merged <- Laag_Inkomen_merged %>%
  mutate(
    working_poor = Totaal_werkende_armen / BevolkingOp1Januari_1
  )
#Afronding percentage
Laag_Inkomen_merged <- Laag_Inkomen_merged %>%
  mutate(
    working_poor = round((Totaal_werkende_armen / BevolkingOp1Januari_1) * 100, 1)
  )



```
```{r}
# Create a named vector for mapping RegioS codes to city names
city_map <- c(
  "GM0268" = "Nijmegen",
  "GM0307" = "Amersfoort",
  "GM0344" = "Utrecht",
  "GM0363" = "Amsterdam",
  "GM0392" = "Haarlem",
  "GM0523" = "Hardinxveld-Giessendam",
  "GM0534" = "Hillegom",
  "GM0599" = "Rotterdam",
  "GM0532" = "Stede Broec"
)

# Add the new City column by matching RegioS values
Laag_Inkomen_merged$City <- city_map[Laag_Inkomen_merged$RegioS]

```
```{r}
#Delete useless columns
Laag_Inkomen_merged <- Laag_Inkomen_merged[, !(names(Laag_Inkomen_merged) %in% c("InkomensgrensHuishouden", "DuurInkomenspositie", "KenmerkenVanHuishoudens"))]



```
```{r}
# Define vectors for each city type
big_cities <- c("Amsterdam", "Rotterdam", "Utrecht")
medium_cities <- c("Amersfoort", "Nijmegen", "Haarlem")
small_cities <- c("Hardinxveld-Giessendam", "Hillegom", "Stede Broec")

# Create the new column using ifelse
Laag_Inkomen_merged$Type_of_city <- ifelse(
  Laag_Inkomen_merged$City %in% big_cities, "big",
  ifelse(
    Laag_Inkomen_merged$City %in% medium_cities, "medium",
    ifelse(
      Laag_Inkomen_merged$City %in% small_cities, "small",
      NA  # for any city not in the lists
    )
  )
)

```
```{r}
Laag_Inkomen_merged$Covid <- ifelse(
  Laag_Inkomen_merged$Perioden %in% c("2018JJ00", "2019JJ00"), "Pre-Covid",
  ifelse(
    Laag_Inkomen_merged$Perioden %in% c("2022JJ00", "2023JJ00"), "Post-Covid",
    NA  # For any other years or missing values
  )
)

```

```{r}
library(ggplot2)

# Explicitly set factor levels to include all city types
Laag_Inkomen_merged$Type_of_city <- factor(
  Laag_Inkomen_merged$Type_of_city,
  levels = c("small", "medium", "big")
)

# Create the boxplot, forcing all levels to appear on the x-axis
ggplot(Laag_Inkomen_merged, aes(x = Type_of_city, y = working_poor)) +
  geom_boxplot(fill = "skyblue", color = "darkblue", show.legend = TRUE) +
  scale_x_discrete(drop = FALSE) +  # important: do not drop unused factor levels
  labs(
    title = "Working poor percentage by City Type",
    x = "Type of City",
    y = "Working poor percentage"
  ) +
  theme_minimal()




```
```{r}
# Set the factor levels so Pre-Covid comes first, Post-Covid second
Laag_Inkomen_merged$Covid <- factor(Laag_Inkomen_merged$Covid, levels = c("Pre-Covid", "Post-Covid"))

library(ggplot2)

# Create the plot with the reordered factor
ggplot(Laag_Inkomen_merged, aes(x = Covid, y = working_poor, fill = Covid)) +
  geom_boxplot(alpha = 0.5) +
  facet_wrap(~ Type_of_city) +
  labs(
    title = "Working poor percentage Before and After Covid by City Type",
    x = "Covid Period",
    y = "Working poor percentage"
  ) +
  theme_minimal()




```
```{r}
library(dplyr)
library(ggplot2)

# Define the big cities you want to focus on
big_cities <- c("Amsterdam", "Utrecht", "Rotterdam")

# Filter the dataframe for big cities only
Laag_Inkomen_filtered <- Laag_Inkomen_merged %>%
  filter(Type_of_city == "big" & City %in% big_cities)

# Ensure Covid factor levels are ordered correctly
Laag_Inkomen_filtered$Covid <- factor(Laag_Inkomen_filtered$Covid, levels = c("Pre-Covid", "Post-Covid"))

# Create the boxplot faceted by City
ggplot(Laag_Inkomen_filtered, aes(x = Covid, y = working_poor, fill = Covid)) +
  geom_boxplot(alpha = 0.6) +
  facet_wrap(~ City) +
  labs(
    title = "Working Poor percentage Before and After Covid in Big Cities",
    x = "Covid Period",
    y = "Working Poor percentage"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Pre-Covid" = "blue", "Post-Covid" = "red"))

```
```{r}
library(dplyr)
library(ggplot2)

# Define the big cities you want to focus on
big_cities <- c("Amsterdam", "Utrecht", "Rotterdam")

# Filter the dataframe for only these big cities
Laag_Inkomen_filtered_no_time <- Laag_Inkomen_merged %>%
  filter(City %in% big_cities)

# Create the boxplot, grouping by City, without splitting by Covid period
ggplot(Laag_Inkomen_filtered_no_time, aes(x = City, y = working_poor, fill = City)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Overall Working poor percentage Distribution in Big Cities",
    subtitle = "(Aggregated Pre- & Post-Covid Data)",
    x = "City",
    y = "Working poor percentage"
  ) +
  theme_minimal() +
  theme(legend.position = "none") # Hide legend as fill is just for city

```
```{r}

library(sf)
gpkg_path <- "cbsgebiedsindelingen2025.gpkg"
gemeente_gegeneraliseerd <- st_read(gpkg_path, layer = "gemeente_gegeneraliseerd")


```

```{r}
library(dplyr)
merged_spatial <- gemeente_gegeneraliseerd %>%
  left_join(Laag_Inkomen_merged, by = c("statnaam" = "City"))
library(ggplot2)
ggplot(merged_spatial) +
  geom_sf(aes(fill = working_poor), color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", na.value = "grey90") +
  labs(
    title = "Zoomed-In Spatial Distribution of Working Poor percentage",
    fill = "Value"
  ) +
  theme_minimal() +
  coord_sf(
    xlim = c(50000, 200000),  
    ylim = c(420000, 500000)   
  )

```
```{r}
# Load necessary library
library(ggplot2)

# Convert 'Perioden' to year by extracting the first four characters
Laag_Inkomen_merged$Year <- as.numeric(substr(Laag_Inkomen_merged$Perioden, 1, 4))

# Group by 'Year' and sum the 'Totaal_werkende_armen' for each year
df_yearly <- aggregate(Totaal_werkende_armen ~ Year, data = Laag_Inkomen_merged, sum)

# Plot the line graph
ggplot(df_yearly, aes(x = Year, y = Totaal_werkende_armen)) +
  geom_line(color = 'blue', size = 1) +
  geom_point(color = 'red', size = 2) +
  labs(title = "Total Working Poor by Year", x = "Year", y = "Total Working Poor") +
  theme_minimal()

```

```{r}
# Load necessary library
library(ggplot2)

# Convert 'Perioden' to year by extracting the first four characters
Laag_Inkomen_merged$Year <- as.numeric(substr(Laag_Inkomen_merged$Perioden, 1, 4))

# Group by 'Year' and calculate the mean of 'werkende_armen_per_capita' for each year
df_yearly_percentage <- aggregate(working_poor ~ Year, data = Laag_Inkomen_merged, mean)

# Plot the line graph
ggplot(df_yearly_percentage, aes(x = Year, y = working_poor)) +
  geom_line(color = 'blue', size = 1) +
  geom_point(color = 'red', size = 2) +
  labs(title = "Total Percentage of Working Poor by Year", x = "Year", y = "Percentage of Working Poor") +
  theme_minimal()


```

